% =============================================================================
% FILE: frontmatter/acknowledgments.tex
% DESCRIPTION: Acknowledgments section
% =============================================================================

\chapter*{ACKNOWLEDGMENTS}
\addcontentsline{toc}{chapter}{ACKNOWLEDGMENTS}
\thispagestyle{plain}

\setlength{\parindent}{0pt}
\setlength{\parskip}{1em}

% Introduction
I would like to express my deepest gratitude to all those who have supported me throughout this research journey and made this thesis possible.

% Thesis Advisor
First and foremost, I extend my sincere appreciation to my thesis advisor, \textbf{[Advisor Name, Ph.D.]}, for their invaluable guidance, patience, and expertise throughout this research. Their insightful feedback and encouragement have been instrumental in shaping this work and developing my skills as a researcher.

% Panel Members
I am grateful to my thesis panel members, \textbf{[Panel Member 1 Name]}, \textbf{[Panel Member 2 Name]}, and \textbf{[Panel Member 3 Name]}, for their constructive criticism and suggestions that significantly improved the quality of this thesis.

% Department and University
My heartfelt thanks go to the Physics Department and the College of Science and Mathematics at the University of Science and Technology of Southern Philippines for providing the academic environment and resources necessary for this research.

% Colleagues and Peers
I wish to thank my fellow physics students and colleagues, especially \textbf{[Names]}, for the stimulating discussions, collaborative spirit, and moral support throughout this challenging yet rewarding journey.

% Technical Support
Special thanks to \textbf{[Names of technical staff or computational facility]}, for providing access to computational resources and technical assistance essential for training the deep learning models in this study.

% Family
To my family, especially my \textbf{[parents/guardians]}, I am eternally grateful for your unwavering love, support, and encouragement. Your sacrifices and belief in my abilities have been my constant source of strength.

% Friends
I also acknowledge my friends \textbf{[Names]} for their companionship and for keeping me grounded during the most stressful moments of this work.

% Financial Support (if applicable)
% This research was supported in part by [Grant/Scholarship Name, if applicable].

% Closing
Finally, I thank the open-source community and the developers of PyTorch, the Swin Transformer, and related libraries that made this research possible.

\vspace{1cm}

\begin{flushright}
\textit{Dean Mark H. Catiil}\\
\textit{November 2025}
\end{flushright}

\cleardoublepage